Real Time Variational Density Forecasting

### 2017-02-21 Confirmation/confirmation_V2.tex
Substantial rewrite to report

%2multibyte Version: 5.50.0.2960 CodePage: 1252
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[centertags]{amsmath}
\usepackage{graphicx}%
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[hidelinks]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{amsmath}
\usepackage[]{algorithm2e}
\usepackage{amsthm}
\usepackage{url}
\usepackage{wasysym}
\usepackage{ulem}
\usepackage{afterpage}
\usepackage{bbm}
\setcounter{MaxMatrixCols}{30}
\providecommand{\U}[1]{\protect\rule{.1in}{.1in}}
\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0.1in}
\setlength{\evensidemargin}{0.1in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{8.25in}
\numberwithin{equation}{section}
\title{Real Time Variational Density Forecasts}
\author{Nathaniel Tomasetti, Catherine Forbes and Anastasios Panagiotelis}
\begin{document}

@@

Bayesian methods are computationally complex as they require integration over the set of unknowns, whether parameters or latent stochastic variables. This set is quite large for models used to model short term electricity load and price as there are strong seasonal patterns in the data \citep{Taylor2003}. While analytical integration is typically not feasible to obtain the predictive distribution for such models, however there are a range of numerical and simulation-based techniques available for this purpose, notably Markov Chain Monte Carlo (MCMC) methods. While it is often possible to demonstrate theoretical convergence of MCMC algorithms, computation may be slow for even moderately complex models. This is problematic for when attempting to forecast electricity load in Victoria, when realizations occur at five minute intervals, yhe predictive density for the next period's electricity load may not be available before it is observed.

@@

The discussion of MCMC and VB will be continued in Section \ref{sec:BayesInf}, with the methods contrasted in the context of a simple time series model. Section \ref{sec:ElecLoad} will introduce a planned emperical application of VB to Victorian electricity load using an exponential smoothing model and data from the 2012-2015 period. Section \ref{sec:Discussion} outlines the major areas this thesis will make a contribution to and Section \ref{sec:Timeline} outlines the planned timeline of these contributions.

@@

To facilitate the discussion, consideration is given to models where the unknowns may be summarized in a (finite) $k-$dimensional vector, denoted by $\boldsymbol{\theta}.$ In this setting, given an observed time series, denoted by $y_{1:T}= \{y_{t}, t = 1, \dots T\}$, the Bayesian forecast distribution associated with time $t=T+1$ is characterized by the conditional density
\begin{equation}
\label{predictive}
p(y_{T+1} | y_{1:T}) =\int p(y_{T+1}|y_{1:T},\boldsymbol{\theta}) p(\boldsymbol{\theta}|y_{1:T}) d\boldsymbol{\theta},
\end{equation}
where $p(\boldsymbol{\theta}|y_{1:T})$ denotes the posterior density for $\boldsymbol{\theta}$, given by
\begin{equation}
\label{posterior}
 p(\boldsymbol{\theta} | y_{1:T}) = \frac{p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta}}{\int_{\boldsymbol{\theta}}p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta}) d\boldsymbol{\theta}}.
\end{equation}

Generally the analytical solution to (\ref{posterior}), and hence to (\ref{predictive}) will be unavailable due to the complex functional form of the distributions involved. 
In this section two alternative methods for computing the desired posterior distribution in (\ref{posterior}),  MCMC and VB, will be reviewed.
In brief, MCMC is used to create a sample from $p(\boldsymbol{\theta} | y_{1:T})$, with any function of $\boldsymbol{\theta}$ that is desired estimated from that sample. In contrast, VB replaces $p(\boldsymbol{\theta} | y_{1:T})$ with an approximation, denoted by $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, where $\boldsymbol{\lambda}$ is a vector of auxiliary parameters associated with the approximation and may depend on the observations $y_{1:T}$.
For example, if $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is Gaussian, then $\boldsymbol{\lambda}$ would denote the mean and variance parameters that characeterise this normal distribution for $\boldsymbol{\theta}$.  Variational Bayes aims to choose the distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ such that the Kullback-Leibler divergence from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to $p(\boldsymbol{\theta} | y_{1:T})$ is minimised. To obtain the predictive distribution in (\ref{predictive}), $y_{T+1}$ can be treated as an unknown variable in $\boldsymbol{\theta}$ and hence directly computed by either MCMC and VB. Alternatively, once the posterior distribution in (\ref{posterior}) is available, (\ref{predictive}) can be found by simulation.

@@

\begin{align}
\mu | \sigma^2, y_{1:T} &\sim \mathcal{N} \left(\frac{\tau\gamma + T\bar{y}}{\tau + T}, \frac{\sigma^2}{\tau  + T} \right) \label{posmu} \\
\sigma^2 | y_{1:T} &\sim Inv.Gamma  \left( \mbox{shape} = \alpha + T/2, \mbox{scale} = \beta + \frac{1}{2}\sum_{t=1}^{T}y_t^2 + \tau\gamma^2 - \frac{(\tau\gamma + T\bar{y})^2}{\tau + T} \right) \label{possig2} 
\end{align}
where $\bar{y} = 1/T \sum_{t=1}^{T}y_t$. Posterior sampling can proceed by first simulating a draw of $\sigma^2$ from (\ref{possig2}) and conditioning on this draw to simulate $\mu$ from (\ref{posmu}).

@@

Since $\ln(p(y_{1:T}))$ is constant with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, maximising (\ref{ELBO}) with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is equivalent to minimising (\ref{KL-def}). Maximising the ELBO is much more convenient than minimising the KL Divergence, as will be shown in the following sections in the context of the two major implementations of Varitaional Bayes, namely Mean Field Variational Bayes and Stochastic Variational Bayes. Each of these implementations take advantage of the functional form of the ELBO, and offer computationally efficient algorithms to find a $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that is optimal within a particular class of distributions. 
of distributions.

@@

MFVB is widely used as it greatly simplifies maximisation of the ELBO, and has a straight-forward implementation in exponential family models \citep{Wainwright2008}.  Maximising the ELBO with respect to $q_i$ is as analytically involved as deriving the conditional distributions used in Gibbs based MCMC schemes, but the resulting algorithm to maximise the ELBO is much simpler computationally than sampling from a Gibbs MCMC scheme. MFVB factorises (\ref{ELBO}) into a product of $k$ seperate components that each contain only one $q_i$, and then applies an algorithm to maximise each of these factors with respect the associated $q_i$. Using the notation $q_{\setminus i} = \prod_{j\neq i}q_j$, \citet{Attias1999} shows that each component is maximised by
\begin{equation}
\label{mf2}
q_i(\theta_i |\boldsymbol{\lambda_i}) \propto\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y_{1:T},\boldsymbol{\theta}))]).
\end{equation}

We illustrate the implementation of MFVB with the model considered in Section (\ref{sec:MCMC}), where $y_t$ for $t = 1, \dots, T$ is generated independently from a $\mathcal{N}(\mu, \sigma^2)$ distribution, and prior distributions are given by $\mu \sim \mathcal{N}(\gamma, \sigma^2 / \tau)$ and $\sigma^2 \sim Inv.Gamma(\mbox{shape} = \alpha, \mbox{scale =}\beta)$. It can be shown that
\begin{equation}
\label{mf3}
q(\mu | \boldsymbol{\lambda}_1) \propto \exp \left\{ \frac{-(T\tau + \mathbb{E}_{q(\sigma^2)}(\sigma^{-2}))}{2\tau\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})} \left( \mu - \frac{\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})\gamma + \lambda \sum_{t=1}^{T} y_t}{T \tau + \mathbb{E}_{q(\sigma^2 )}(\sigma^{-2})} \right)^2 \right\}
\end{equation}
and
\begin{equation}
\label{mf4}
q(\sigma^2 | \boldsymbol{\lambda}_2) \propto \sigma^{-2(T/2 + \alpha + 1)} \exp \left\{ \frac{ -1/2(\sum_{t=1}^{T}y_t^2 + n\mathbb{E}_{q(\mu)}(\mu^2) - 2\sum_{t=1}^{T} y_t \mathbb{E}_{q(\mu)}(\mu)) - \beta}{\sigma^2} \right\}.
\end{equation}

As the prior distributions used were conjugate to the likelihood, the Variational Bayes optimal distributions for $q(\mu |  \boldsymbol{\lambda}_1)$ and $q(\sigma^2 | \boldsymbol{\lambda}_2)$ are the same family as the prior distributions, and are denoted by $q(\mu | \boldsymbol{\lambda}_1) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\tau})$ and $q(\sigma^2 | \boldsymbol{\lambda}_2) \sim Inv.Gamma(\tilde{\alpha}, \tilde{\beta})$. The vector $\boldsymbol{\lambda} = (\tilde{\gamma}, \tilde{\tau}, \tilde{\alpha}, \tilde{\beta})'$ is given by
\begin{align}
\tilde{\gamma} &= \frac{\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})\gamma + \lambda \sum_{t=1}^{T} y_t}{T \tau + \mathbb{E}_{q(\sigma^2 )}(\sigma^{-2})}  \label{mf5} \\ 
\tilde{\tau} &= \frac{\tau\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})}{(T\tau + \mathbb{E}_{q(\sigma^2)}(\sigma^{-2}))} \label{mf6} \\
\tilde{\alpha} &= T/2 + \alpha  \label{mf7} \\
\tilde{\beta} &= 1/2\left(\sum_{t=1}^{T}y_t^2 + n\mathbb{E}_{q(\mu)}(\mu^2) - 2\sum_{t=1}^{T} y_t \mathbb{E}_{q(\mu)}(\mu))\right) + \beta. \label{mf8}
\end{align}.

The known values of the expectations in (\ref{mf5}) - (\ref{mf8}) for these distributions can be substituted into these expressions to derive the set of mean field equations:

\begin{align}
\tilde{\gamma} &= \frac{\tilde{\alpha} / \tilde{\beta} \gamma + \tau \sum_{t=1}^{T} y_t} {T \lambda + \tilde{\alpha} / \tilde{\beta}} \label{mf9} \\ 
\tilde{\tau} &= \frac{T \tilde{\alpha} / \tilde{\beta}}{T \tau + \tilde{\alpha} / \tilde{\beta}} \label{mf10} \\
\tilde{\beta} &= \frac{1}{2} \left(\sum_{t=1}^{T} y_t^2 + T(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \sum_{t=1}^{T} y_t \tilde{\gamma} \right) + \beta. \label{mf11}
\end{align}
$\tilde{\alpha}$ is omitted from the mean field equations as its value does not depend on other $\boldsymbol{\lambda}$ parameters. (\ref{mf9}) - (\ref{mf11}) contain a circular dependence in these equations an algorithm that iteratively cycles through these equations until the change in the ELBO is less than some pre-specified constant $\epsilon$. This algorithm is known as a coordiant ascent algorithm and is described by Algorithm (\ref{alg:algorithm1}), at each step it will fix all but one parameter and select the value of the omitted parameter which maximises the ELBO.

The set of mean field equations are only obtainable in the event that a prior that is conjugate to the model likelihood is used, restricting MFVB to exponential family models. In this case the optimal distributional family $q_i$ is the same as the prior, analagous to the relationship between a prior-posterior conjugate pair. In the event that a non-conjugate prior is used, the form of the distribution in (\ref{mf2}) is unrecognisable, a further approximation may be used by substituting in another distribution to replace the problematic $q_i(\theta_i |\boldsymbol{\lambda_i}).$ A common option is the use of a Gaussian distribution through either a Laplace approximation \citep{Friston2006}, a Delta Method inspired transform \citet{Wang2013}, though the only requirement is that the substitute distribution has moments that are tractable functions of its parameters to be used for other parameters in (\ref{mf2}). This secondary level of approximations is analogous to the requirement of a Metropolis-Hastings-within-Gibbs step in MCMC to handle unrecognisable full conditional distributions.

\vspace{5mm}
\begin{algorithm}[H]
 \SetKwInOut{Input}{Input}
 \Input{Log Joint Density}
 \KwResult{Mean Field Approximation}
 Use (\ref{mf3}) to match $q(\mu|\boldsymbol{\lambda}_1)$ to a Gaussian distributon and $q(\sigma^2 | \boldsymbol{\lambda}_2)$ to an Inverse Gamma distribution\;
 Anaytically derive the set of mean field equations in \ref{mf9}) - (\ref{mf11})\;
 Initialise $\boldsymbol{\lambda}^{(1)}$ randomly.\;
 Evaluate $\mathcal{L}(q, y_{1:T})^{(1)}$ using $\boldsymbol{\lambda}^{(1)}$.\;
 Set $\tilde{\alpha}$ using \ref{mf7}.\;
 \While{$\mathcal{L}(q, y_{1:T})^{(m)} - \mathcal{L}(q, y_{1:T})^{(m-1)} > \epsilon$}{
  Set $\tilde{\gamma}^{(m)}$ using \ref{mf9}.\;
  Set $\tilde{\tau}^{(m)}$ using \ref{mf10}.\;
  Set $\tilde{\beta}^{(m)}$ using \ref{m11}.\;
  Evaluate $\mathcal{L}(q, y_{1:T})^{(m)}$.\;
  Set $m = m + 1$.\;
 }
 \caption{Coordinate Ascent for MFVB}
  \label{alg:algorithm1}
\end{algorithm}

@@

The requirement for an exponential family model, as well as restricting the approximating distribution to a factorisable family may be unsatisfactory in many applications where the dependence between parameters is important, as in the AR(2) model where Figure \ref{fig:MCMCplot} shows strong dependence between $\phi_1$ and $\phi_2$. Without factorisation the simple method to maximise the ELBO as in MFVB is unavailable. \citet{Paisley2012} and \citet{Ranganath2014} have adapted a gradient ascent algorithm for use in Variational Bayes, resulting in what is known as Stochastic Variational Bayes (SVB).

The algorithm for SVB iteratively takes the  derivative of $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ in (\ref{ELBO}) with respect to $\boldsymbol{\lambda}$ and the following updating step is applied until the ELBO converges to a maximum.

\begin{equation}
\label{SGA1}
\boldsymbol{\lambda}^{(m+1)} = \boldsymbol{\lambda}^{(m)} + \rho^{(m)} \nabla_{\boldsymbol{\lambda}} \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}),
\end{equation}

where $\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ is the vector of partial derivatives of $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ with respect to each element of $\boldsymbol{\lambda}$. This update requires some initial values $\boldsymbol{\lambda}^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate. If $\rho^{(m)}$ is chosen to satisfy the following conditions the algorithm is guaranteed to converge to a local maximum \citep{Robbins1951}.

\begin{align}
&\sum_{m=1}^{\infty} \rho^{(m)} =  \infty \\
&\sum_{m=1}^{\infty} (\rho^{(m)})^2 <  \infty.
\end{align}

>>>>

%2multibyte Version: 5.50.0.2960 CodePage: 1252
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[centertags]{amsmath}
\usepackage{graphicx}%
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[hidelinks]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{amsmath}
\usepackage[]{algorithm2e}
\usepackage{amsthm}
\usepackage{url}
\usepackage{wasysym}
\usepackage{ulem}
\usepackage{tikz}
%\usetikzlibrary{bayesnet}
\usepackage{afterpage}
\usepackage{bbm}
\setcounter{MaxMatrixCols}{30}
\providecommand{\U}[1]{\protect\rule{.1in}{.1in}}
\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0.1in}
\setlength{\evensidemargin}{0.1in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{8.25in}
\numberwithin{equation}{section}
\title{Real Time Variational Density Forecasting}
\author{Nathaniel Tomasetti, Catherine Forbes and Anastasios Panagiotelis}
\begin{document}

@@

Bayesian methods are computationally complex as they require integration over the set of unknowns, whether parameters or latent stochastic variables. This set is quite large for models used to model short term electricity load and price as there are strong seasonal patterns in the data \citep{Taylor2003}. While analytical integration is typically not feasible to obtain the predictive distribution for such models, however there are a range of numerical and simulation-based techniques available for this purpose, notably Markov Chain Monte Carlo (MCMC) methods. While it is often possible to demonstrate theoretical convergence of MCMC algorithms, computation may be slow for even moderately complex models. This is problematic for when attempting to forecast electricity load in Victoria, when realizations occur at five minute intervals, the predictive density for the next period's electricity load may not be available before it is observed.

@@

The discussion of MCMC and VB will be continued in Section \ref{sec:BayesInf}, with the methods contrasted in the context of two simple models. Section \ref{sec:ElecLoad} will introduce a planned emperical application of VB to Victorian electricity load using an exponential smoothing model and data from the 2012-2015 period. Section \ref{sec:Discussion} outlines the major areas to which this thesis will make a contribution, and Section \ref{sec:Timeline} outlines the planned timeline for these contributions.

@@

To facilitate the discussion, consideration is given to models where the unknowns may be summarized in a (finite) $k-$dimensional vector, denoted by $\boldsymbol{\theta}.$ In this setting, given an observed time series, denoted by $y_{1:T}= \{y_{t}, t = 1, \dots T\}$, the Bayesian forecast distribution associated with time $t=T+1$ is characterized by the conditional density
\begin{equation}
\label{predictive}
p(y_{T+1} | y_{1:T}) =\int p(y_{T+1}|y_{1:T},\boldsymbol{\theta}) p(\boldsymbol{\theta}|y_{1:T}) d\boldsymbol{\theta},
\end{equation}
where $p(\boldsymbol{\theta}|y_{1:T})$ denotes the posterior density for $\boldsymbol{\theta}$, given by
\begin{equation}
\label{posterior}
 p(\boldsymbol{\theta} | y_{1:T}) = \frac{p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta})}{\int_{\boldsymbol{\theta}}p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta}) d\boldsymbol{\theta}}.
\end{equation}

Generally the analytical solution to (\ref{posterior}), and hence to (\ref{predictive}) will be unavailable due to the complex functional form of the distributions involved. 
In this section two alternative methods for computing the desired posterior distribution in (\ref{posterior}),  MCMC and VB, will be reviewed.
In brief, MCMC is used to create a sample from $p(\boldsymbol{\theta} | y_{1:T})$, with any function of $\boldsymbol{\theta}$ that is desired estimated from that sample. In contrast, VB replaces $p(\boldsymbol{\theta} | y_{1:T})$ with an approximation, denoted by $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, where $\boldsymbol{\lambda}$ is a vector of auxiliary parameters associated with the approximation and may depend on the observations $y_{1:T}$.
For example, if $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is Gaussian, then $\boldsymbol{\lambda}$ would denote the mean and variance parameters that characeterise this normal distribution for $\boldsymbol{\theta}$.  Variational Bayes aims to choose the distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ such that the Kullback-Leibler divergence from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to $p(\boldsymbol{\theta} | y_{1:T})$ is minimised. Once the posterior distribution in (\ref{posterior}) is available, the predictive density forecast (\ref{predictive}) can be found by simulation.

@@

\begin{align}
&\mu | \sigma^2, y_{1:T} \sim \mathcal{N} \left(\frac{\tau\gamma + T\bar{y}}{\tau + T}, \frac{\sigma^2}{\tau  + T} \right) \label{posmu} \\
&\sigma^2 | y_{1:T} \sim Inv.Gamma  \left( \mbox{shape} = \alpha + T/2, \mbox{scale} = \beta + \frac{1}{2}\sum_{t=1}^{T}y_t^2 + \tau\gamma^2 - \frac{(\tau\gamma + T\bar{y})^2}{\tau + T} \right) \label{possig2} 
\end{align}
where $\bar{y} = 1/T \sum_{t=1}^{T}y_t$. Posterior sampling can proceed by first simulating $\sigma^2$ from (\ref{possig2}) and conditioning on this draw to simulate $\mu$ from (\ref{posmu}). It should be noted that as we are simulating from $p(\sigma^2 | y)$ rather than $p(\sigma^2 | \mu, y)$, there is no Markov dependency between iterations and this is a Monte Carlo sampler rather than an MCMC.

@@

Since $\ln(p(y_{1:T}))$ is constant with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, maximising (\ref{ELBO}) with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is equivalent to minimising (\ref{KL-def}). Maximising the ELBO is much more convenient than minimising the KL Divergence, as will be shown in the following sections in the context of the two major implementations of Varitaional Bayes, namely Mean Field Variational Bayes and Stochastic Variational Bayes. Each of these implementations take advantage of the functional form of the ELBO, and offer computationally efficient algorithms to find a $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that is optimal within a particular class of distributions.

@@

MFVB is widely used as it greatly simplifies maximisation of the ELBO, and has a straight-forward implementation in exponential family models \citep{Wainwright2008}.  Maximising the ELBO with respect to $q_i$ is as analytically involved as deriving the conditional distributions used in Gibbs based MCMC schemes, but the resulting algorithm to maximise the ELBO is much simpler computationally than sampling from a Gibbs MCMC scheme. MFVB factorises (\ref{ELBO}) into a product of $k$ seperate components that each contain only one $q_i$, and then maximises each of these factors with respect to the associated $q_i$. Using the notation $q_{\setminus i} = \prod_{j\neq i}q_j$, \citet{Attias1999} shows that each component is maximised by
\begin{equation}
\label{mf2}
q_i(\theta_i |\boldsymbol{\lambda_i}) \propto\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y_{1:T},\boldsymbol{\theta}))]).
\end{equation}

We illustrate the implementation of MFVB with the model considered in Section (\ref{sec:MCMC}), where $y_t$ for $t = 1, \dots, T$ is generated independently from a $\mathcal{N}(\mu, \sigma^2)$ distribution, and prior distributions are given by $\mu \sim \mathcal{N}(\gamma, \sigma^2 / \tau)$ and $\sigma^2 \sim Inv.Gamma(\mbox{shape} = \alpha, \mbox{scale =}\beta)$. It can be shown that
\begin{equation}
\label{mf3}
q(\mu | \boldsymbol{\lambda}_1) \propto \exp \left\{ \frac{-(T + \tau) \mathbb{E}_{q(\sigma^2)}(\sigma^{-2})}{2} \left( \mu - \frac{\tau \gamma + \sum_{t=1}^{T} y_t}{T + \tau} \right)^2 \right\}
\end{equation}
and
\begin{align}
\label{mf4}
q(\sigma^2 | \boldsymbol{\lambda}_2) &\propto \sigma^{-2((T+1)/2 + \alpha + 1)} \nonumber \\
\times &\exp \left\{ \frac{ \beta + 1/2\left((\tau + T)\mathbb{E}_{q(\mu)} - 2 \mathbb{E}_{q(\mu)}(\mu)(\sum_{t=1}^{T}y_t + \tau) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2\right)}{-\sigma^2} \right\}.
\end{align}

As the prior distributions used were conjugate to the likelihood, the Variational Bayes optimal distributions for $q(\mu |  \boldsymbol{\lambda}_1)$ and $q(\sigma^2 | \boldsymbol{\lambda}_2)$ are the same family as the prior distributions, and are denoted by $q(\mu | \boldsymbol{\lambda}_1) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\tau})$ and $q(\sigma^2 | \boldsymbol{\lambda}_2) \sim Inv.Gamma(\tilde{\alpha}, \tilde{\beta})$. The vector $\boldsymbol{\lambda} = (\tilde{\gamma}, \tilde{\tau}, \tilde{\alpha}, \tilde{\beta})'$ is given by
\begin{align}
\tilde{\gamma} &= \frac{\tau \gamma + \sum_{t=1}^{T} y_t}{T + \tau}  \label{mf5} \\ 
\tilde{\tau} &= \left((T + \tau)\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})\right )^{-1} \label{mf6} \\
\tilde{\alpha} &= (T+1)/2 + \alpha  \label{mf7} \\
\tilde{\beta} &= \beta + 1/2\left((\tau + T)\mathbb{E}_{q(\mu)} - 2 \mathbb{E}_{q(\mu)}(\mu)(\sum_{t=1}^{T}y_t + \tau) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right). \label{mf8}
\end{align}.

The known values of the expectations in (\ref{mf5}) - (\ref{mf8}) for these distributions can be substituted into these expressions to derive the set of mean field equations:
\begin{align}
\tilde{\tau} &= \frac{\tilde{\beta}}{\tilde{\alpha}(T + \tau)} \label{mf9} \\
\tilde{\beta} &= \beta + 1/2\left((\tau + T)(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \tilde{\gamma}(\sum_{t=1}^{T}y_t + \tau) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right)\label{mf10}
\end{align}
$\tilde{\gamma}$ and $\tilde{\alpha}$ are omitted from the mean field equations as their values do not depend on other $\boldsymbol{\lambda}$ parameters.

As (\ref{mf9}) and (\ref{mf10}) contain a circular dependence, fitting the MFVB approximation requires an algorithm that cycles between these equations until the change in the ELBO is less than some pre-specified small constant $\epsilon$. This algorithm is known as a coordiant ascent, and is described by Algorithm (\ref{alg:algorithm1}). At each step coordinate ascent will fix one parameter and then maximise the ELBO with respect to the other parameter, so that changes in parameters results in a non-decreasing sequence of ELBO values. Coordinate ascent converges only when neither parameter can be changed independently to increase the ELBO by more than $\epsilon$, and hence reaches a local maximum.
The current theory for Variational Bayes does not offer any insight to the shape of the ELBO so determining if the algorithm converges to a local or global maximum is difficult. One option to alleiviate this problem is to start the algorithm at a range of initial values and choose parameter values with the maximum converged ELBO, and hence lowest KL divergence to the true posterior.

Figure (\ref{fig:norminvg}) compares the simulated posterior densities from Section (\ref{sec:MCMC}), (red) with the approximation from Mean Field Variational Bayes (blue) and the analytically available true posterior marginal densities (black). For both parameters all three densities are indistinguishable.

\begin{figure}[h]
\centering
\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{norminvg}
\caption{Marginal posterior densities for the model described in Section (\ref{sec:MCMC}). The true density is in black, the Monte Carlo simulation is red and the MFVB approximation is blue. Both MCMC and MFVB fit this density extremely well.}
\label{fig:norminvg}
\end{figure}

The set of mean field equations are only obtainable in the event that a prior that is conjugate to the model likelihood is used, restricting MFVB to exponential family models. In this case the optimal distributional family $q_i$ is the same as the prior, analagous to the relationship between a prior-posterior conjugate pair. In the event that a non-conjugate prior is used, the form of the distribution in (\ref{mf2}) is unrecognisable, a further approximation may be used by substituting in another distribution to replace the problematic $q_i(\theta_i |\boldsymbol{\lambda_i}).$ A common option is the use of a Gaussian distribution through either a Laplace approximation \citep{Friston2006} or a Delta Method inspired transform \citet{Wang2013}; though the only requirement is that the substitute distribution has moments that are tractable functions of its parameters to be used for other parameters in (\ref{mf2}). This secondary level of approximations is analogous to the requirement of a Metropolis-Hastings-within-Gibbs step in MCMC to handle unrecognisable full conditional distributions.

\vspace{5mm}
\begin{algorithm}[H]
 \SetKwInOut{Input}{Input}
 \Input{Log Joint Density}
 \KwResult{Mean Field Approximation}
 Use (\ref{mf2}) to match $q(\mu|\boldsymbol{\lambda}_1)$ to a Gaussian distributon and $q(\sigma^2 | \boldsymbol{\lambda}_2)$ to an Inverse Gamma distribution\;
 Anaytically derive the set of mean field equations in (\ref{mf9}) - (\ref{mf10})\;
 Initialise $\boldsymbol{\lambda}^{(1)}$ randomly.\;
 Evaluate $\mathcal{L}(q, y_{1:T})^{(1)}$ using $\boldsymbol{\lambda}^{(1)}$.\;
 Set $\tilde{\gamma}$ using \ref{mf5}.\;
 Set $\tilde{\alpha}$ using \ref{mf7}.\;
 \While{$\mathcal{L}(q, y_{1:T})^{(m)} - \mathcal{L}(q, y_{1:T})^{(m-1)} > \epsilon$}{
  Set $\tilde{\tau}^{(m)}$ using \ref{mf9}.\;
  Set $\tilde{\beta}^{(m)}$ using \ref{mf10}.\;
  Evaluate $\mathcal{L}(q, y_{1:T})^{(m)}$.\;
  Set $m = m + 1$.\;
 }
 \caption{Coordinate Ascent for MFVB}
  \label{alg:algorithm1}
\end{algorithm}

@@

The requirement for an exponential family model without introducing further approximations, as well as restricting the approximating distribution to a factorisable family may be unsatisfactory in many applications. \citet{Paisley2012} and \citet{Ranganath2014} have adapted a gradient ascent algorithm for use in Variational Bayes, which can maximise the ELBO for non-factorisable functional forms of $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$. We refer to the result as Stochastic Variational Bayes (SVB).

The application of gradient ascent in SVB iteratively takes the derivative of $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ in (\ref{ELBO}) with respect to $\boldsymbol{\lambda}$ and the following updating step is applied until change in the ELBO is less than $\epsilon$.
\begin{equation}
\label{SVB1}
\boldsymbol{\lambda}^{(m+1)} = \boldsymbol{\lambda}^{(m)} + \rho^{(m)} \nabla_{\boldsymbol{\lambda}} \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}),
\end{equation}
where $\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ is the vector of partial derivatives of $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ with respect to each element of $\boldsymbol{\lambda}$. This update requires some initial values $\boldsymbol{\lambda}^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate. If $\rho^{(m)}$ is chosen to satisfy the following conditions the algorithm is guaranteed to converge to a local maximum \citep{Robbins1951}.
\begin{align}
&\sum_{m=1}^{\infty} \rho^{(m)} =  \infty \\
&\sum_{m=1}^{\infty} (\rho^{(m)})^2 <  \infty.
\end{align}
